Fisher vector places: learning compact descriptors for place recognition

### 2014-11-13 cvpr15review.tex
polish + gitignore

      \textcolor{petr}{%TODO: \emph{we should't speak about calibration, we shoud address not ignoring $b$}\\
         Notice that even the new descriptor $\Phi_j$ is obtained by normalization of the learnt hyperplane $w_j$, the bias $b$ form equation~\eqref{eq:decomposed} is not ignored. Indeed, the bias term in convex objective~\eqref{eq:obj} is affecting the learnt the hyperplane.
         %\\
         %The proposed re-normalization procedure can be interpreted as a simple form of affine calibration where the uncalibrated SVM score $s(x)=w^T x+b$ is transformed into a calibrated score $f(x)=\alpha s(x)+\beta$, where $\alpha = 1/||w||$ and $\beta = -b\alpha$. Note that the bias term b is not ignored but used to compute $f(x)$. The intuition is that for L2 normalized descriptors x computing the calibrated score $f(x)$ corresponds to computing the normalized cross-correlation between vectors w and x. This was found to work well in retrieval (see e.g. Sivic and Zisserman, ICCV 2003 [32]) or matching whitened HOG descriptors [11,13] (see e.g. [Doersch et al., Mid-Level Visual Element Discovery as discriminative Mode Seeking, NIPS 2013]), but has not yet been used in conjunction with SVMs as done in this work. Note also that the affine calibration can be computed offline and does not need any additional computation or storage at query time compared to the p-value calibration of Gronat et al. [12].
      }

@@

   \subsection{Results}  
      \textcolor{petr}{TODO: comment results from the table - raw SVM BOW/FV}  
      For each database (Pittsburgh 25k and Pittsburgh 55k) we compare results of our method (FV e-SVM) to two baselines: standard bag-of-visual-words baseline (BOW) and raw Fisher vector matching without learning (FV).

      We perform experiments on several target Fisher vector dimensions $d\in\{128,512,2048\}$. For each method we measure performance using the percentage of correctly recognized queries (Recall) similarly to, e.g.,~\cite{Chen11,Knopp2010,Sattler-BMVC12}. The query is correctly localized if at least one of the top $K$ retrieved database images is within $20$ meters from the ground truth position of the query. Results are shown for different values of $K$ in table~\ref{tab:recall}. For the Pittsburgh 25k we also show results in the form of a curve in figure~\ref{fig:recall}. The results clearly demonstrate the benefits of the learnt descriptors with respect to the standard Fisher vectors for all target dimensions and lengths of shortlist $K$. The benefits of discriminative learning are specially prominent for low-dimensional compact descriptors ($d=128$). 
      The proposed method also significantly outperforms the bag-of-visual-words baseline. Figure~\ref{fig:images} shows examples of place recognition results.

      \paragraph{Applying the method to other descriptors.}
        \textcolor{petr}{ 
          The proposed approach can be applied to other descriptors beyond Fisher vectors. To demonstrate this we have applied the proposed method to the bag-of-visual-words descriptor and observed improvement from 28.7\% recall@K=1 (vanilla bag-of-visual-words) to 31.8\% recall@K=1 (e-SVM). For bag-of-visual-words, the e-SVM with p-value calibration gives comparable results (33.6\% recall@K=1) but at an increased training cost and a significantly higher required memory footprint at query time, which makes the method very hard to scale-up to database sizes beyond 200k images.
        }

      \paragraph{Comparison to other methods.}
         On the Pittsburgh 25k database, we compare performance of our learnt discriminative descriptors to the methods of~\cite{Gronat13} and~\cite{Knopp2010}, who report on the same testing data top $K=1$ recall of 36.5\% and 41.9\%, respectively (results taken from~\cite{Gronat13}). Our method outperforms~\cite{Knopp2010} already for dimension $d=128$ (37.8\%) and~\cite{Gronat13} for dimension $d=512$ (47.6\%). Furthermore, note that~\cite{Gronat13,Knopp2010} are based on a bag-of-visual-words representation, which typically needs to store between 1000-2000 non-zero visual words per image, which is significantly more than our learnt 128 or 512 dimensional descriptor.

        \textcolor{petr}{
         We also performed p-val calibration proposed by Gronat et al.~\cite{Gronat13} on e-SVM Fisher vectors. We found that the calibration did not perform for the learnt Fisher vector classifiers (top 1 recall of 25.3\% compared to baseline performance of 33.6\% for dimension 128). When examining the results, we believe this may be attributed to the fact that Fisher vectors are low dimensional compared to the extremely high-dimensional (d=100,000) bag-of-visual-words representation, which affects the distribution of the classifier scores on the negative test data. 
          %We have compared the proposed method with the p-value calibration of (Gronat et al.[12]) applied to Fisher vectors. The results show that our method significantly outperforms the p-value calibration (37.8\% recall@1 vs. 32.3\% recall@1; 56.9\%recall@5 vs. 50.0\% recall@5) for the Fisher vector of dimension 128. Similar gains are observed for other FV dimensions.
        }
      \paragraph{Memory complexity analysis.}
         Figure~\ref{fig:memory} compares the performance of the learnt discriminative descriptor (FV eSVM) to the raw Fisher vectors (FV) for different target dimensions. 
         The results demonstrate that for a given level of accuracy (y-axis) our method learns a more compact (lower-dimensional) representation (x-axis). For example, our learnt 128-dimensional descriptor achieves a similar accuracy (around 65\%) to the 256-dimensional raw Fisher descriptor essentially reducing the memory complexity to 50\% for the same level of performance.  Note that 
         similar to~\cite{Jegou12}, we observe decrease in performance at high-dimensions for both the baseline and our method.
         %
         \begin{figure}[t!]
            \centering
            \includegraphics[width=1.0\linewidth]{imgs/FVmemory}    
            \caption{
               \textbf{Memory complexity analysis.} 
               The fraction of correctly localized queries at the top 10 retrieved images (y-axis) for different Fisher vector dimensions (x-axis). The learnt descriptors by our method (FV e-SVM) clearly outperform the raw Fisher vector descriptors (FV) for all dimensions. Note that for a certain level of performance (y-axis) the proposed method learns a more memory efficient (lower dimensional, x-axis) descriptor.
            }
            \label{fig:memory}
         \end{figure}
         \begin{table}[htb]
            \begin{centering}
               \include{table25k}
            \caption{ \textbf{Results on Pittsburgh 25k dataset.}
                  The fraction of correctly recognized queries (recall@K)
                  \textcolor{petr}{
                     in top $K\in\{1,2,5,10,20\}$ retrieved images.
                     Different methods have been applied to two types of descriptors, the BOW and Fisher vectors compressed to different dimensions.The learnt representations (BOW \emph{e-SVM} and BOW \emph{p-val}) outperform the raw BOW baseline as well as the learnt representation without calibration (BOW raw SVM).
                  } 
                  The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors across the whole range of $K$  and all dimensions on both the 25k and 55k Pittsburgh image datasets.     
            }
            \label{tab:recall}
            \end{centering}
         \end{table}

         \begin{table}[htb]
            \begin{centering}
               \include{table55k}
               \caption{ \textbf{Results on Pittsburgh 55k dataset.}
                  The fraction of correctly recognized queries (recall@K)
                  \textcolor{petr}{
                  in top $K\in\{1,2,5,10,20\}$ retrieved images.
                  The proposed method (\emph{e-SVM}) has been applied to Fisher vectors compressed to different dimensions.
                  } 
                  The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors across the whole range of $K$  and all dimensions on both the 25k and 55k Pittsburgh image datasets.     
               }
            \label{tab:recall2}
            \end{centering}
         \end{table}
         %
         %
         %
         \begin{figure*}[h]
            \include{myFigureWnorm}
            \caption{
                \textbf{Examples of correctly and incorrectly localized queries for the learnt Fisher vector representation.}
                Each example shows a query image (left) together with correct (green) and incorrect (red) matches from the database obtained by the learnt Fisher vector representation \emph{w-norm} method (top) and the standard Fisher vector baseline (bottom) for dimension 128. Note that the proposed method is able to recognize the place depicted in the query image despite changes in viewpoint, illumination and partial occlusion by other objects (trees, lamps) and buildings. 
                Note that the baseline methods often finds images depicting the same buildings but in a distance whereas our learnt representation often finds a closer view better matching the content of the query.       
            }
            \label{fig:images}
         \end{figure*}
         %
         %
         %

>>>>

      \textcolor{petr}{%TODO: \emph{we should't speak about calibration, we shoud address not ignoring $b$}\\
         The intuition is that for L2 normalized query descriptors $\Phi_q$ computing the calibrated score $f(x)$ corresponds to computing the normalized cross-correlation between vectors $\Psi_j$ and $\Phi_q$. This was found to work well in retrieval (see e.g.~\cite{Sivic03}) or matching whitened HOG descriptors \cite{Gharbi12, Hariharan12}, but has not yet been used in conjunction with SVMs. 
         \\
         Notice that even the new descriptor $\Psi_j$ is obtained by normalization of the learnt hyperplane $w_j$, the bias $b$ form equation~\eqref{eq:decomposed} is not ignored. Indeed, the bias term in convex objective~\eqref{eq:obj} is affecting the learnt the hyperplane.
         %\\
         %The proposed re-normalization procedure can be interpreted as a simple form of affine calibration where the uncalibrated SVM score $s(x)=w^T x+b$ is transformed into a calibrated score $f(x)=\alpha s(x)+\beta$, where $\alpha = 1/||w||$ and $\beta = -b\alpha$. Note that the bias term b is not ignored but used to compute $f(x)$. The intuition is that for L2 normalized descriptors x computing the calibrated score $f(x)$ corresponds to computing the normalized cross-correlation between vectors w and x. This was found to work well in retrieval (see e.g. Sivic and Zisserman, ICCV 2003 [32]) or matching whitened HOG descriptors [11,13] (see e.g. [Doersch et al., Mid-Level Visual Element Discovery as discriminative Mode Seeking, NIPS 2013]), but has not yet been used in conjunction with SVMs as done in this work. Note also that the affine calibration can be computed offline and does not need any additional computation or storage at query time compared to the p-value calibration of Gronat et al. [12].
      }

@@

   \subsection{Results}  
      For each database (Pittsburgh 25k and Pittsburgh 55k) we compare results of our method (FV e-SVM) to two baselines: standard bag-of-visual-words baseline (BOW) and raw Fisher vector matching without learning (FV).

      We perform experiments on several target Fisher vector dimensions $d\in\{128,512,2048\}$. For each method we measure performance using the percentage of correctly recognized queries (Recall) similarly to, e.g.,~\cite{Chen11,Knopp2010,Sattler-BMVC12}. The query is correctly localized if at least one of the top $K$ retrieved database images is within $20$ meters from the ground truth position of the query. Results are shown for different values of $K$ in tables~\ref{tab:recall} and \ref{tab:recall2}. For the Pittsburgh 25k we also show results in the form of a curve in figure~\ref{fig:recall}. The results clearly demonstrate the benefits of the learnt descriptors with respect to the standard Fisher vectors for all target dimensions and lengths of shortlist $K$. The benefits of discriminative learning are specially prominent for low-dimensional compact descriptors ($d=128$). 
      The proposed method also significantly outperforms the bag-of-visual-words baseline. Figure~\ref{fig:images} shows examples of place recognition results.

      \paragraph{Applying our learning method to other descriptors.}
        \textcolor{petr}{ 
         The proposed approach can be applied to other descriptors beyond Fisher vectors. To demonstrate this we have applied the proposed method to the bag-of-visual-words descriptor and observed improvement from 28.7\% recall@K=1 (vanilla bag-of-visual-words) to 31.8\% recall@K=1 (e-SVM). For bag-of-visual-words, the e-SVM with p-value calibration gives comparable results (33.6\% recall@K=1) but at an increased training cost and a significantly higher required memory footprint at query time, which makes the method very hard to scale-up to database sizes beyond 200k images. The results are given in table \ref{tab:recall}.
        }

      \paragraph{Comparison to other methods.}
         %On the Pittsburgh 25k database, we compare performance of our learnt discriminative descriptors to the methods of~\cite{Gronat13} and~\cite{Knopp2010}, who report on the same testing data top $K=1$ recall of 36.5\% and 41.9\%, respectively (results taken from~\cite{Gronat13}). Our method outperforms~\cite{Knopp2010} already for dimension $d=128$ (37.8\%) and~\cite{Gronat13} for dimension $d=512$ (47.6\%). Furthermore, note that~\cite{Gronat13,Knopp2010} are based on a bag-of-visual-words representation, which typically needs to store between 1000-2000 non-zero visual words per image, which is significantly more than our learnt 128 or 512 dimensional descriptor. 
         \textcolor{petr}{
         On the Pittsburgh 25k database, we compare performance of our learnt discriminative descriptors to the method of~\cite{Gronat13} who report on the similar testing data. Our method outperforms~\cite{Gronat13} already for dimension $d=128$ (37.8\%).
         \\
         The p-val calibration proposed by Gronat et al.~\cite{Gronat13} was also tested on e-SVM Fisher vectors. We found that the calibration did not perform for the learnt Fisher vector classifiers (top 1 recall of 25.3\% compared to baseline performance of 33.6\% for dimension 128). When examining the results, we believe this may be attributed to the fact that Fisher vectors are low dimensional compared to the extremely high-dimensional (d=100,000) bag-of-visual-words representation, which affects the distribution of the classifier scores on the negative test data.
         \\
         Finally, we investigate how performs a raw exemplar-SVM without any further calibration both image descriptors . We observe that while using raw SVM scores directly without any calibration fails on BOW (as reported in \cite{Gronat13}), it marginally improves the recall for larger $K$ for Fisher vectors as indicated in \ref{tab:recall}.
          %We have compared the proposed method with the p-value calibration of (Gronat et al.[12]) applied to Fisher vectors. The results show that our method significantly outperforms the p-value calibration (37.8\% recall@1 vs. 32.3\% recall@1; 56.9\%recall@5 vs. 50.0\% recall@5) for the Fisher vector of dimension 128. Similar gains are observed for other FV dimensions.
        }
      \paragraph{Memory complexity analysis.}
         Figure~\ref{fig:memory} compares the performance of the learnt discriminative descriptor (FV eSVM) to the raw Fisher vectors (FV) for different target dimensions. 
         The results demonstrate that for a given level of accuracy (y-axis) our method learns a more compact (lower-dimensional) representation (x-axis). For example, our learnt 128-dimensional descriptor achieves a similar accuracy (around 65\%) to the 256-dimensional raw Fisher descriptor essentially reducing the memory complexity to 50\% for the same level of performance.  Note that 
         similar to~\cite{Jegou12}, we observe decrease in performance at high-dimensions for both the baseline and our method.
         %
         \begin{figure}[t!]
            \centering
            \includegraphics[width=1.0\linewidth]{imgs/FVmemory}    
            \caption{
               \textbf{Memory complexity analysis.} 
               The fraction of correctly localized queries at the top 10 retrieved images (y-axis) for different Fisher vector dimensions (x-axis). The learnt descriptors by our method (FV e-SVM) clearly outperform the raw Fisher vector descriptors (FV) for all dimensions. Note that for a certain level of performance (y-axis) the proposed method learns a more memory efficient (lower dimensional, x-axis) descriptor.
            }
            \label{fig:memory}
         \end{figure}
         \begin{table}[b]
            \begin{centering}
               \include{table25k}
            \caption{ \textbf{Results on Pittsburgh 25k dataset.}
                  The fraction of correctly recognized queries (recall@K)
                  \textcolor{petr}{
                     in top $K\in\{1,2,5,10,20\}$ retrieved images.
                     Different methods have been applied to two types of descriptors, the BOW and Fisher vectors compressed to different dimensions.The learnt representations (BOW \emph{e-SVM} and BOW \emph{p-val}) outperform the raw BOW baseline as well as the learnt representation without calibration (BOW raw SVM).
                  } 
                  The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors across the whole range of $K$  and all dimensions on both the 25k and 55k Pittsburgh image datasets.     
            }
            \label{tab:recall}
            \end{centering}
         \end{table}

         \begin{table}[t]
            \begin{centering}
               \include{table55k}
               \caption{ \textbf{Results on Pittsburgh 55k dataset.}
                  The fraction of correctly recognized queries (recall@K)
                  \textcolor{petr}{
                  in top $K\in\{1,2,5,10,20\}$ retrieved images.
                  The proposed method (\emph{e-SVM}) has been applied to Fisher vectors compressed to different dimensions.
                  } 
                  The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors across the whole range of $K$  and all dimensions on both the 25k and 55k Pittsburgh image datasets.     
               }
            \label{tab:recall2}
            \end{centering}
         \end{table}
         %
         %
         %
         \begin{figure*}[h]
            \include{myFigureWnorm}
            \caption{
                \textbf{Examples of correctly and incorrectly localized queries for the learnt Fisher vector representation.}
                Each example shows a query image (left) together with correct (green) and incorrect (red) matches from the database obtained by the learnt Fisher vector representation \emph{e-SVM} method (top) and the standard Fisher vector baseline (bottom) for dimension 128. Note that the proposed method is able to recognize the place depicted in the query image despite changes in viewpoint, illumination and partial occlusion by other objects (trees, lamps) and buildings. 
                Note that the baseline methods often finds images depicting the same buildings but in a distance whereas our learnt representation often finds a closer view better matching the content of the query.       
            }
            \label{fig:images}
         \end{figure*}
         %
         %
         %
